---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r echo=FALSE, message=FALSE}
set.seed(2020)
require(dirichletprocess)
require(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/img/"
)
```

# dirichletprocess

[![Travis-CI Build Status](https://travis-ci.org/dm13450/dirichletprocess.svg?branch=master)](https://travis-ci.org/dm13450/dirichletprocess)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/dm13450/dirichletprocess?branch=master&svg=true)](https://ci.appveyor.com/project/dm13450/dirichletprocess)
[![Coverage Status](https://codecov.io/gh/dm13450/dirichletprocess/branch/master/graph/badge.svg)](https://codecov.io/gh/dm13450/dirichletprocess)

The dirichletprocess package provides tools for you to build custom Dirichlet process mixture models. You can use the pre-built Normal/Weibull/Beta distributions or create your own following the instructions in the vignette. In as little as four lines of code you can be modelling your data nonparametrically.

## Installation

You can install the stable release of dirichletprocess from CRAN:

```{r, eval=FALSE}
install.packages("dirichletprocess")
```

You can also install the development build of dirichletprocess from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("dm13450/dirichletprocess")
```

For a full guide to the package and its capabilities please consult the vignette:

```{r vignette, eval=FALSE}
browseVignettes(package = "dirichletprocess")
```

## Examples

### Density Estimation

Dirichlet processes can be used for nonparametric density estimation. 

```{r density, eval=TRUE}
faithfulTransformed <- faithful$waiting - mean(faithful$waiting)
faithfulTransformed <- faithfulTransformed/sd(faithful$waiting)
dp <- DirichletProcessGaussian(faithfulTransformed)
dp <- Fit(dp, 100, progressBar = FALSE)
plot(dp)
```

```{r}
data.frame(Weight=dp$weights, Mean=c(dp$clusterParameters[[1]]), SD=c(dp$clusterParameters[[1]]))
```


### Clustering

Dirichlet processes can also be used to cluster data based on their common distribution parameters. 

```{r clustering, eval=TRUE}
faithfulTrans <- as.matrix(apply(faithful, 2, function(x) (x-mean(x))/sd(x)))
dpCluster <-  DirichletProcessMvnormal(faithfulTrans)
dpCluster <- Fit(dpCluster, 1000, progressBar = FALSE)
```

To plot the results we take the cluster labels contained in the `dp` object and assign them a colour

```{r clusteringplot, echo=FALSE}
faithRes <- cbind(faithful, Labels = dpCluster$clusterLabels)
ggplot(faithRes, aes(x=eruptions, y=waiting, colour=as.factor(Labels))) + geom_point() + guides(colour=FALSE)
```

For more detailed explanations and examples see the vignette.


