---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# dirichletprocess

The dirichletprocess package provides tools for you to build custom Dirichlet Process mixture models. You can use the prebuilt Normal/Weibull/Beta distribtuions or create your own following the instructions in the vignette. 

## Installation

You can install dirichletprocess from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("dm13450/dirichletprocess")
```

For a full guide to the package and its capabilites please consult the vignette:

```{r vignette, eval=FALSE}
browseVignettes(package = "dirichletprocess")
```

## Example

### Density Estimation

```{r example, eval=FALSE}
faithfulTransformed <- faithful$waiting - mean(faithful$waiting)
faithfulTransformed <- faithfulTransformed/sd(faithful$waiting)
dp <- DirichletProcessGaussian(faithfulTransformed)
dp <- Fit(dp, 100, progressBar = FALSE)
plot(dp)
```


### Clustering

```{r clustering, eval=FALSE}
faithfulTrans <- as.matrix(apply(faithful, 2, function(x) (x-mean(x))/sd(x)))
dp <-  DirichletProcessMvnormal(faithfulTrans)
dp <- Fit(dp, 100, progressBar = FALSE)
```




